---
title: "MAM"
author: "Elisabet Cueli Alba"
date: "2025-12-02"
output: pdf_document
---

```{r}
library(fpp2)
library(forecast)
data(ausbeer)
ausbeer
```
**1. Use ets {forecast} to estimate ausbeer {fpp2}**
```{r}
ausbeer_ets <- ets(ausbeer)
ausbeer_ets
```
**2. Use ets (model = "AAA") {forecast} to estimate ausbeer {fpp2}**
```{r}
ausbeer_ets_aaa <- ets(model="AAA", ausbeer)
ausbeer_ets_aaa
```
**3. Compare the accuracy {forecast} of both models in-sample**
```{r}
rbind(accuracy(ausbeer_ets), accuracy (ausbeer_ets_aaa))
```
**4. Plot the 1-to-20-step-ahead forecasts for the first model**
```{r}
forecast_ets <- forecast(ausbeer_ets, h = 20)
plot(forecast_ets, main = "1-to-20-Step-Ahead Forecasts")
```
**5. Cross Validation (CV): Use an increasing rolling training set, starting with the first 180 observations, to fit the first model, and make one-step-ahead forecasts. Find the one-step-ahead forecast errors**
```{r}
n <- length(ausbeer)#total number of observation
m <- 4 #always 4 quarters

H <- n - 180 #number of periods to forecast
forecast_errors <- numeric(H) #vector for forecast errors

for (i in 1:H) { #loop for each h in H
  train <- window(ausbeer, end = time(ausbeer)[180+i-m])
  forecast_cv <- forecast(ets(train), h = m) #forecast of the next quarter
  forecast_errors[i] <- ausbeer[180 + i] - forecast_cv$mean[m] #we match the quarters
}
```
**6. Plot in one screen the one-step-ahead forecast errors and the squared one-step-ahead forecast errors**
```{r}
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))

plot(forecast_errors, type = "l", main = "One-Step-Ahead Forecast Errors")

plot(forecast_errors^2, type = "l", main = "Squared One-Step-Ahead Forecast Errors")

par(mfrow = c(1, 1))
```
**7. Find the one-step-ahead out-of-sample RMSFE**
```{r}
rmsfe <- sqrt(mean(forecast_errors^2))
print(paste("One-Step-Ahead Out-of-Sample RMSFE:", round(rmsfe, 2)))
```

