---
title: "Seasonal ETS"
author: "Elisabet Cueli Alba"
date: "2025-11-28"
output: pdf_document
---

```{r}
library(fpp2)
library(forecast)
data(austourists)
```
**1. Estimate austourists using ets()**

```{r}
austourists_ets <- ets(austourists)
summary(austourists_ets)
```
**2. Plot in one screen the data together with the level, slope, and seasonal components**

```{r}
par(mfrow = c(4, 1), mar = c(3, 4, 2, 1))

plot(austourists, main = "Original Data")

plot(austourists_ets$states[, "l"], main = "Level")
plot(austourists_ets$states[, "b"], main = "Slope")
plot(austourists_ets$states[, "s1"], main = "Seasonal")

par(mfrow = c(1, 1))
```

**3. Plot one-to-eight-ahead forecasts**
```{r}
austourists_ets_forecast8 <- forecast(austourists_ets,h=8)
plot(austourists_ets_forecast8,main="Forecast one-to-eight ahead")
```
**4. Use hw {forecast} to estimate austourists {fpp2}, starting at 2010, with additive and multiplicative seasonality**

*Additive*
```{r}
hw_additive <- hw(window(austourists, start=2010), seasonal = "additive", h = 8)
summary(hw_additive)
```
*Multiplicative*
```{r}
hw_multiplicative <- hw(window(austourists, start=2010), seasonal = "multiplicative", h = 8)
summary(hw_multiplicative)
```
**5. Plot in one screen the data together with the within-sample and out-of-sample one-step-ahead forecasts of the additive and multiplicative models**
```{r}
plot(austourists, main = "Additive and Multiplicative Forecasts", type="o", 
     xlim = c(2010, 2018), ylim = c(20, 70))

lines(window(austourists, start=2010), col = "black")

lines(fitted(hw_additive), col = "blue", lwd = 2)
lines(fitted(hw_multiplicative), col = "red")

lines(hw_additive$mean, col = "purple")
lines(hw_multiplicative$mean, col = "pink")

# Add legend
legend("bottomright", 
       legend = c("Data", "Additive Fitted", "Multiplicative Fitted", 
                  "Additive Forecast", "Multiplicative Forecast"),
       col = c("black", "blue", "red", "purple", "pink"),
       lty = c(1, 1, 1, 2, 2),
       lwd = 2, cex = 0.8)
```
**6. Plot in one screen the level, slope, and seasonal components of the additive and multiplicative models**
```{r}
par(mfrow = c(3, 2))

plot(hw_additive$model$states[, "l"], main = "Level of Additive", ylab="")
plot(hw_multiplicative$model$states[, "l"], main = "Level of Multiplicative", ylab="")

plot(hw_additive$model$states[, "b"], main = "Slope of Additive", ylab="")
plot(hw_multiplicative$model$states[, "b"], main = "Slope of Multiplicative", ylab="")

plot(hw_additive$model$states[, "s1"], main = "Seasonal of Additive", ylab="")
plot(hw_multiplicative$model$states[, "s1"], main = "Seasonal of Multiplicative", ylab="")
par(mfrow = c(1, 1))
```

