---
title: "SSARMA (1,2) - Financial Econometrics"
author: "Elisabet Cueli Alba"
date: "2025-11-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(FKF)
set.seed(123)
```
**1. Generate Synthetic ARMA(1,2) Data**

We generate 1000 observations from an ARMA(1,2) process with parameters:
$\phi_1 = 0.7$, $\theta_1 = 0.4$, $\theta_2 = 0.3$, $\sigma = \sqrt2$

```{r}
# True parameters
ar1 <- 0.7
ma1 <- 0.4
ma2 <- 0.3
sigma <- sqrt(2)

# Generate ARMA(1,2) data
n <- 1000
arma_model <- list(ar = ar1, ma = c(ma1, ma2))
y <- arima.sim(model = arma_model, n = n, innov = sigma*rnorm(n))

# Plot the series
plot(y, main = "Synthetic data 1000 ARMA(1,2)")
grid()
```
**2. State Equation**

The ARMA(1,2) model can be written as:
$$Y_t = \phi_1 Y_{t-1} + W_t + \theta_1 W_{t-1} + \theta_2 W_{t-2}$$
The **state equation** is:
$$\alpha_t = T_t \alpha_{t-1} + \sigma_W \eta_t$$

Where:
$$T_t = \begin{bmatrix} \phi_1 & \phi_2=0 & \phi_3=0 \\ 1 & 0 & 0 \\ 0 & 1 & 0 \end{bmatrix}, \quad \sigma_W \eta_t= \begin{bmatrix} W_t \\ 0 \\ 0 \end{bmatrix}$$
Thus, the state equation is of this form:

$$\begin{bmatrix} \alpha_{t,1} \\ \alpha_{t,2} \\ \alpha_{t,3} \end{bmatrix} = \,
\begin{bmatrix} \phi_1 & 0 & 0 \\ 1 & 0 & 0 \\ 0 & 1 & 0 \end{bmatrix} \,
\begin{bmatrix} \alpha_{t-1,1} \\ \alpha_{t-1,2} \\ \alpha_{t-1,3} \end{bmatrix} + \,
\begin{bmatrix} W_t \\ 0 \\ 0 \end{bmatrix}$$


**3. Observation Equation**

Letting $r \equiv \max\{p, q+1\}$, r=3 in our case (ARMA(1,2)). Given no constant and $G_t=0$, the **observation equation** is:
$$Y_t = C_t + Z_t \alpha_t + G_t \epsilon_t = Z_t \alpha_t = Z_t \begin{bmatrix} \alpha_{t,1} \\ \alpha_{t,2} \\ \alpha_{t,3} \end{bmatrix}$$

Where:
$$Z_t = \begin{bmatrix} 1 & \theta_1 & ... & \theta_{r-1} \end{bmatrix} = \begin{bmatrix} 1 & \theta_1 & \theta_2 \end{bmatrix} $$


```{r}
```
**4. Function arma12ss()**

This function creates the state-space representation matrices for an ARMA(1,2) model.

```{r}
arma12ss <- function(ar1, ma1, ma2, sigma) {
  ## initial value of the state variable:
  r <- 3
  a0 <- c(0, 0, 0)
  P0 <- matrix(1e6, nrow=r, ncol=r) # MSE (diffuse prior)
  
  ## State equation:
  dt <- matrix(0, nrow = r)
  Tt <- matrix(c(ar1,0,0, 1,0,0, 0,1,0), ncol=r) # t(T) TRANSPOSE
  Ht <- sigma * matrix(c(1, 0, 0, 0, 0, 0, 0, 0, 0), ncol=r)
  HHt <- Ht %*% t(Ht)
  
  ## Observation equation:
  ct <- matrix(0)
  Zt <- matrix(c(1, ma1, ma2), ncol=3)
  GGt <- matrix(0)
  
  return(list(a0 = a0, P0 = P0,
             dt = dt, ct = ct,
             Tt = Tt, Zt = Zt,
             HHt = HHt, GGt = GGt))
 }
```
**5. Objective Function for Maximum Likelihood**

We create a function that returns the negative log-likelihood to be minimized.

```{r}
objective <- function(param,yt){
  #SS representation of ARMA(1,2)
  ss <- arma12ss(
    ar1=param["ar1"],
    ma1=param["ma1"],
    ma2=param["ma2"],
    sigma=param["sigma"])
  
  ans <- fkf(a0 = ss$a0, P0 = ss$P0, #output of arma12 is input for fkf
             dt = ss$dt, ct = ss$ct,
             Tt = ss$Tt, Zt = ss$Zt,
             HHt = ss$HHt, GGt = ss$GGt,
             yt = yt)
  
  return(-ans$logLik)
}
```
**6. Parameter Estimation using optim()**

Estimation is then performed using a numeric search by optim {stats}.

```{r}
param0 <- c(ar1=0, ma1=0, ma2=0, sigma=1)

 fit <- optim(par = param0, # initial values
              fn = objective, # logLik to be maximized
              yt = rbind(y), # (1,1000) to be passed to "fn"
              hessian = TRUE)
names(fit)
```
```{r}
fit$par
```
**7. Construct 95% Confidence Intervals**

We compute standard errors from the Hessian matrix and construct 95% confidence intervals for each parameter.
```{r}
p <- cbind(
  actual=c(ar1=ar1, ma1=ma1, ma2=ma2, sigma=sigma),
  estimate=fit$par,
  lowerCI = fit$par - qnorm(0.975)*sqrt(diag(solve(fit$hessian))),
  upperCI = fit$par + qnorm(0.975)*sqrt(diag(solve(fit$hessian)))
)
p
```
**Conclusion**

Only actual ar1 and sigma seems to fall out of the Confidence Interval, but the CI for ma1 and ma2 seem statistical unsignificant as they are too wide.

$\,$

**8. Filter the Series using fkf()**

We apply the Kalman filter with the estimated parameters to obtain filtered state estimates.
```{r}
sp<- arma12ss(
  fit$par["ar1"], fit$par["ma1"],fit$par["ma2"], fit$par["sigma"]
)

names(sp)
```
```{r}
ans <- fkf(
  a0 = sp$a0, 
  P0 = sp$P0,
  dt = sp$dt, 
  ct = sp$ct,
  Tt = sp$Tt, 
  Zt = sp$Zt,
  HHt = sp$HHt, 
  GGt = sp$GGt,
  yt = rbind(y)
)
names(ans)
```
**9. Plot Predicted Series vs Realization**
```{r}
plot(ans,at.idx=1,att.idx=NA, CI=NA)
lines(as.vector(y), lty ="dotted")
```
**10. Plot Updated Series vs Realization**
```{r}
plot(ans,at.idx=NA,att.idx=1, CI=NA)
lines(as.vector(y), lty ="dotted")
```
**11. Diagnostic Plots: QQ and ACF of Residuals**

We examine the residuals to verify the normality and independence assumptions.
```{r}
plot(ans,type="resid.qq")
```
**12. Smoothed Estimates using fks()**

The Kalman smoother produces $a_{t|n}$, which uses all available information from the entire sample.

```{r}
sm <- fks(ans)
names(sm)
```
```{r}
plot(sm,CI=NA,ahatt.idx = 1)
lines(as.vector(y),lty="dotted")

```
**Conclusion**

All forecasted, updated and smoothed series are very closed to the realized series, denoting a good fit for the model. 

Regarding the Q-Q graphic, the body of the distribution is well alike to the Normal Distribution, though with a slight deviation in the tails being heavier, meaning higher probabilities for both negative and positive outliers, characteristic that resambles a T-Student distribution.